Disk Winds Matter: Simulating the Optical Spectra of Cataclysmic Variables

### 2014-08-20 draft.tex
Some small changes to typos, abstracts and inclusion of more notes

\begin{abstract}
High-State non-magnetic Cataclysmic Variables (CVs) 
exhibit strong Hydrogen \& Helium recombination lines in their optical 
spectra. These lines can be single peaked, and may be formed 
in the same bipolar outflow that is responsible for the
blue-shifted absorption lines seen in the Ultraviolet (UV).
Here we present results obtained by incorporating a `macro-atom' treatment into
our Monte Carlo radiative transfer code, which was
originally used to synthesize the UV spectrum of a disk-dominated CV 
with a simple biconical wind. Using an identical model, we show that the wind leaves a noticeable imprint 
on the optical spectrum, producing He II 4686 and H${\alpha}$ as well as
enhanced emission in the Ly-$\alpha$ and He II $1640$ UV lines.
In addition, the improved treatment of recombination means that the bound-free continuum
emission is enhanced, albeit by a modest amount.
A short exploration of a 3-variable kinematic parameter space leads to models with higher densities 
that are able to produce single-peaked line emission, and also generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge. 
Finally, we present a synthetic spectrum of a next generation model in and out of eclipse,
and identify the strengths and shortcomings of the model in the context of the high inclination
Nova-like variable RW Tri.
\end{abstract}

@@

Outflows are
close to ubiquitous  in accreting systems across many order of magnitude in mass,
from young stellar objects \citep{bunn1995} right up 
to quasi-stellar objects \citep[QSOs;][]{ganguly2008}. Outflows can take the form of jets, such as those
seen in X-ray binaries \citep{fender2006} or active galactic nuclei \citep[AGN;][]{marscher2006}, but can also manifest as `winds' emanating from
the accretion disk.
Understanding the impact of accretion disk winds on spectral features has far-reaching implications. 
They provide a feedback mechanism to limit the growth of black holes in AGN \citep{silkrees1998},
and their ever-present nature implies a deep connection with the accretion process. This connection
is important to understand in order to probe the universality of accretion \citep{mchardy2006, SKUK2012}, 
and to determine the true extent to which
accretion disks conform to the `$\alpha$-disk' model proposed by 
\cite{shakurasunyaev1973}. 
Despite this, in many types of systems the impact of winds on observational
features is somewhat unknown, and many theoretical questions relating to their origin 
and driving mechanism remain unanswered.

@@

While the effect of the wind on the UV is at least partially clear, it may also have more subtle 
effects across the spectral range. It has been proposed that an accretion disk wind
can cause single-peaked emission line profiles in the optical (Murray \& Chiang 1996), 
such as those seen in Nova-like variables and SW Sex stars (Honeycutt, Schlegel,
\& Kaitchuck 1986; Dhillon \& Rutten 1995). It is particularly telling that single-peaked 
lines can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can be seen in P-Cygni like signatures,
observed in lines such as H$_{\alpha}$ and He \textsc{i} $\lambda5876$ \citep{RN98}.
In addition, \cite{KLWB98} suggested
that recombination continuum emission from a disk wind or atmosphere
could be responsible for filling in the Balmer absorption edge. This provides
a potential explanation for
the absence of a pronounced `Balmer jump' in optical spectra of CVs, which has never been adequately produced 
by theoretical models or spectral synthesis.
This dense disk atmosphere or wind transition region
necessary to produce the required recombination emission
could also explain the Fe~\textsc{ii} `curtain'
seen in near-UV observations of the high inclination dwarf nova OY Car in queiscence 
\citep{horne1994}.

@@

Here we present work using the same Monte Carlo radiative transfer code as LK02, but incorporating additional 
`macro-atom' techniques suggested by Lucy (2002, 2003) to improve our treatment of the Hydrogen and Helium
optical lines, as well as recombination continuum emission. 
In section 2 we present the code
and provide a brief description of macro-atoms and the radiative transfer methods we employ. 
We describe the kinematics and geometry of our biconical wind model in section 3.
In section 4 we present a benchmark model, which is discussed in section 5 with particular attention 
devoted to a small-scale parameter search.
Finally, in section 6, we summarise our findings, and outline areas for future work.

@@

@@

We hope that future simulations, coupled with a comparison to data, will 
actually address some of these problems, particularly since
the filtering effect of the wind may cause noticeable effects
on the colour of the emergent spectrum \citep{hassall}. 
Here we use a blackbody illuminating spectrum
for the ionization cycles (see figure~\ref{sed}) and to compute our MC estimators.
During the spectral synthesis stage of the simulation we use stellar atmosphere models.
This is partly because they provide the most pessimistic prediction in terms of whether recombination emission in the 
wind can fill in the Balmer jump, due to the strong absorption beyond the Balmer edge.
For low temperatures, we use stellar atmospheres computed 
by Kurucz (reference), and for $T>50,000$~K we use models computed with 
\textsc{TLUSTY} \citep{tlusty}. The input spectra are created from these model
atmospheres using \textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}.
%The temperature profile of our disk,
%with $\dot{M}_{acc} = 10^{-8}~M_{\odot}yr^{-1}$ can be seen in figure~\ref{t_profile}.

@@

%Whether we include one, why why not, referring to He II and ionization.

@@

\subsection{Ionization and Temperature Structure}

@@

@@

The amount of line and recombination continuum
emission produced by given ion, $i$, in a plasma of volume $V$
depends on the emission measure, $EM$, defined as

@@

@@

\begin{itemize}
	\item producing single peaked line profiles is difficult to say the least
	\item with our high density models we can produce some single peaked line emission,
	mostly in Halpha and beta. the other lines seem to stay double peaked.
	\item this higher density causes (too much) collisionally excited line emission from CIV
	\item if the wind was more equatorial then this might help, but this isn't really supported by observations.
	\item we can definitely reproduce the behaviour predicted by Murray and Chiang
	\item Comment on the observed Balmer decrement being fairly flat, as is observed. Point out that
	this is necessary if H alpha is to be single peaked, as you need significant optical depths.
\end{itemize}
}

@@

We introduce boundary layers with temperatures of 80,000K and 200,000K,
commonly proposed values in the literature. We find that

@@

@@

\subsection{UV Resonance Lines}
Compare spectra to LK02 and other models. Are the spectral features preserved, has anything significant changed.
Lyman alpha?

@@

@@

\begin{itemize}
	\item The depth of the Balmer absorption edge decreases with increasing inclination,
or alternatively that the Balmer recombination emission
is enhanced at higher inclinations, {\sl relative to the disk continuum}.
	\item The He~\textsc{ii}~$\lambda3202{\rm \AA}$ line should be
prominent in cases where He~\textsc{ii}~$\lambda1640{\rm \AA}$ \& 
\heiiopt are seen. 
	\item A Balmer jump in emission during eclipse.
\end{itemize}

@@

>>>>

\begin{abstract}
High-State non-magnetic Cataclysmic Variables (CVs) 
exhibit strong Hydrogen \& Helium recombination lines in their optical 
spectra. These lines can be single peaked, and may be formed 
in the same bipolar outflow that is responsible for the
blue-shifted absorption lines seen in the Ultraviolet (UV).
Here we present results obtained by incorporating a `macro-atom' treatment into
our Monte Carlo radiative transfer code, which was
originally used to synthesize the UV spectrum of a disk-dominated CV 
with a simple biconical wind. Using an identical model, we show that the wind leaves a noticeable imprint 
on the optical spectrum, producing He II 4686 and H${\alpha}$ as well as
enhanced emission in the Ly-$\alpha$ and He II $1640$ UV lines. We find that 
a boundary layer is not essential to produce significant recombination to 
He II, and if one is present we favour temperatures of $<10^5$K to avoid
overionizing the UV resonance line species such as CIV. We also conduct Zanstra method
calculations and estimate that one would infer a boundary layer temperature of around 80,000K
from our benchmark model, which does not include one. This suggests that previous calculations
done using this method are strongly dependent on both the assumed geometry and disk
spectrum.
We also note that
the presence of a boundary layer becomes more noticeable in lower accretion rate systems.
In addition, the improved treatment of recombination means that the bound-free continuum
emission is enhanced, albeit by a modest amount.
A short exploration of a 3-variable kinematic parameter space leads to models with higher densities 
that are able to produce single-peaked line emission, and also generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge.
Finally, we present a synthetic spectrum of a next generation model in and out of eclipse,
and identify the strengths and shortcomings of the model in the context of the high inclination
Nova-like variable RW Tri.
\end{abstract}

@@

Outflows are
close to ubiquitous  in accreting systems across many orders of magnitude in mass,
from young stellar objects \citep{bunn1995} right up 
to quasi-stellar objects \citep[QSOs;][]{ganguly2008}. Outflows can take the form of jets, such as those
seen in X-ray binaries \citep{fender2006} or active galactic nuclei \citep[AGN;][]{marscher2006}, but can also manifest as `winds' emanating from
the accretion disk.
Understanding the impact of accretion disk winds on spectral features has far-reaching implications. 
They provide a feedback mechanism to limit the growth of black holes in AGN \citep{silkrees1998},
and their ever-present nature implies a deep connection with the accretion process. This connection
is important to understand in order to probe the universality of accretion \citep{mchardy2006, SKUK2012}, 
and to determine the true extent to which
accretion disks conform to the `$\alpha$-disk' model proposed by 
\cite{shakurasunyaev1973}. 
Despite this, in many types of systems the impact of winds on observational
features is somewhat unknown, and many theoretical questions relating to their origin 
and driving mechanism remain unanswered.

@@

While the effect of the wind on the UV is at least partially clear, it may also have more subtle 
effects across the spectral range. It has been proposed that an accretion disk wind
can cause single-peaked emission line profiles in the optical (Murray \& Chiang 1996), 
such as those seen in Nova-like variables and SW Sex stars (Honeycutt, Schlegel,
\& Kaitchuck 1986; Dhillon \& Rutten 1995). It is particularly telling that single-peaked 
lines can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can be seen in P-Cygni like signatures,
observed in lines such as \ha and He \textsc{i} $\lambda5876$ \citep{RN98}.
In addition, \cite{KLWB98} suggested
that recombination continuum emission from a disk wind or atmosphere
could be responsible for filling in the Balmer absorption edge. This provides
a potential explanation for
the absence of a pronounced `Balmer jump' in optical spectra of CVs, which has never been adequately produced 
by theoretical models or spectral synthesis.
This dense disk atmosphere or wind transition region
necessary to produce the required recombination emission
could also explain the Fe~\textsc{ii} `curtain'
seen in near-UV observations of the high inclination dwarf nova OY Car in queiscence 
\citep{horne1994}.

@@

Here we present work using the same Monte Carlo radiative transfer code as LK02, \py, but incorporating additional 
`macro-atom' techniques suggested by Lucy (2002, 2003) to improve our treatment of the Hydrogen and Helium
optical lines, as well as recombination continuum emission. 
In section 2 we present the code
and provide a brief description of macro-atoms and the radiative transfer methods we employ. 
We describe the kinematics and geometry of our biconical wind model in section 3.
In section 4 we present a benchmark model, which is discussed in section 5 with particular attention 
devoted to a small-scale parameter search. In section 6 we present a `next generation'
model which better reproduces the spectral features observed in NLs, and 
finally, in section 7, we summarise our findings.

% {\bf
% \noindent Fairly happy with this as an introduction. Could maybe
% discuss the boundary layer background here instead of in section 2.
% }

@@

% \noindent\rule{16cm}{0.4pt}

% {\bf
% \noindent Notes on Section 2:

% \begin{itemize}
% 	\item Should this section go after section 3, so that we've got the reader hooked with the model description first?
% 	\item 
% 	\item Would appreciate feedback on whether explanation of macro-atoms 
% 	is sufficient.

% \noindent Potential Additional Figures:

% \begin{itemize}
% 	\item Include figure on Case A/ Case B and TARDIS comparison	
% }
%
% \noindent\rule{16cm}{0.4pt}

@@

We hope that future simulations, coupled with a comparison to data, will 
actually address some of these problems, particularly since
the filtering effect of the wind may cause noticeable effects
on the colour of the emergent spectrum \citep{hassall}. 
Here we use a blackbody illuminating spectrum
for the ionization cycles (see figure~\ref{sed}) and to compute our MC estimators.
During the spectral synthesis stage of the simulation we use stellar atmosphere models.
This is partly because they provide the most pessimistic prediction in terms of whether recombination emission in the 
wind can fill in the Balmer jump, due to their intrinsic strong absorption beyond the Balmer edge.
For low temperatures, we use stellar atmospheres computed 
by Kurucz \citep{kurucz1991}, and for $T>50,000$~K we use models computed with 
\textsc{TLUSTY} \citep{tlusty}. The input spectra are created from these model
atmospheres using \textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}.
%The temperature profile of our disk,
%with $\dot{M}_{acc} = 10^{-8}~M_{\odot}yr^{-1}$ can be seen in figure~\ref{t_profile}.
% more detail on why we use this hodge podge of atmosphere models?

@@

% \noindent\rule{16cm}{0.4pt}
%
% {\bf
% \noindent Notes on Section 3:

% \begin{itemize}
% 	\item

% \noindent Potential Additional Figures:

% \begin{itemize}
% 	\item 	
% }
%
% \noindent\rule{16cm}{0.4pt}

@@

%\subsection{Ionization and Temperature Structure}

@@

% \noindent\rule{16cm}{0.4pt}

% {\bf
% \noindent Notes on Section 4:

% \begin{itemize}
% 	\item

% \noindent Potential Additional Figures:

% \begin{itemize}
% 	\item 	
% }

% \noindent\rule{16cm}{0.4pt}

@@

The amount of line and recombination continuum
emission produced by a given ion, $i$, in a plasma of volume $V$
depends on the emission measure, $EM$, defined as

@@

%% Should I quote Optical depths or 'Q' here to motivate the parameter search?

@@

\begin{itemize}
	\item producing single peaked line profiles is difficult to say the least
	\item with our high density models we can produce some single peaked line emission,
	mostly in Halpha and beta. the other lines seem to stay double peaked.
	\item this higher density causes (too much) collisionally excited line emission from CIV
	\item if the wind was more equatorial then this might help, but this isn't really supported by observations.
	\item we can definitely reproduce the behaviour predicted by Murray and Chiang
	\item Comment on the observed Balmer decrement being fairly flat, as is observed. Point out that
	this is necessary if H alpha is to be single peaked, as you need significant optical depths.
	\item \la needs to be discussed somewhere- potentially here. Personally I would like to point out
	that recombination being important here suggests that it might also be in QSOs
\end{itemize}
}

@@

We introduce boundary layers with temperatures of 80,000K and 200,000K,
commonly proposed values in the literature. We find that a boundary layer

@@

% Possible figure: CIV line equivalent width or ionization fraction
% as a function of boundary layer temperature?

@@

% \noindent\rule{16cm}{0.4pt}

% {\bf
% \noindent Notes on Section 5:

% \begin{itemize}
% 	\item

% \noindent Potential Additional Figures:

% \begin{itemize}
% 	\item 	
% }

% \noindent\rule{16cm}{0.4pt}

% \subsection{UV Resonance Lines}
% Compare spectra to LK02 and other models. Are the spectral features preserved, has anything significant changed.
% Lyman alpha?

@@

% \noindent\rule{16cm}{0.4pt}

% {\bf
% \noindent Notes on Section 6:

% \begin{itemize}
% 	\item

% \noindent Potential Additional Figures:

% \begin{itemize}
% 	\item 	
% }

% \noindent\rule{16cm}{0.4pt}

@@

\begin{itemize}
	\item The depth of the Balmer absorption edge decreases with increasing inclination,
or alternatively that the Balmer recombination emission
is enhanced at higher inclinations, {\sl relative to the disk continuum}.
	\item The He~\textsc{ii}~$\lambda3202{\rm \AA}$ line should be
prominent in cases where He~\textsc{ii}~$\lambda1640{\rm \AA}$ \& 
\heiiopt\ are seen. 
	\item A Balmer jump in emission during eclipse.
\end{itemize}

@@

% {\bf
% \noindent Notes on Section 7:

% \begin{itemize}
% 	\item

% \noindent Potential Additional Figures:

% \begin{itemize}
% 	\item 	
% }
